MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-04-30 MAESTROeX_I/equations.tex
paper updates

\section{Governing Equations}
Low Mach number models for reacting flow were originally derived using asymptotic analysis
\citep{rehm1978equations,majda1985derivation} and used in terrestrial combustion applications
\citep{knio1999semi,day2000numerical}.  These models have been extended to nuclear flames
in astrophysical environments using adaptive algorithms in space and time \citep{Bell:2004}.
In Papers I-III, we derived a model and algorithm suitable for stratified astrophysical flow.
In our model, we take the standard equations of reacting, compressible flow, and recast the equation
of state (EOS) as a divergence constraint on the velocity field.
The resulting model is a series of evolution equations for mass, momentum, and energy, subject
to this constraint:
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc,\label{eq:enthalpy}
\end{eqnarray}
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).\label{eq:U divergence}
\end{equation}
Here $\rho$, $\Ub$, and $h$ are the mass density,
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$.  The species are constrained
such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
Here $\Hnuc$ is the nuclear energy generation rate per unit mass.
The pressure is decomposed into a hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that 
$|\pi|/p_0 = \mathcal{O}(M^2)$ (we use $\xb$ to represent the Cartesian coordinate 
directions of the full state and $r$ to represent the radial coordinate direction for 
the base state).  We also define a base state density, $\rho_0 = \rho_0(r,t)$, 
which is in hydrostatic equilibrium with $p_0$, i.e., 
$\nabla p_0 = -\rho_0 g\eb_r$, where $g=g(r,t)$ is
the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the
outward radial direction.

Mathematically, equations (\ref{eq:momentum})-(\ref{eq:enthalpy}) must still be closed by an equation of state.  
This is done by taking the Langrangian derivative of the equation of state for pressure as a function of the density, composition, and enthalpy, requiring that the pressure is a prescribed function of altitutde and time based on the hydrostatic equilibrium condition.
Details are given in Papers I and II.
After substuting the equations for mass and energy into the Lagrangian derivative, the result
is a divergence constraint on the velocity field (\ref{eq:U divergence}),
where $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho)$ at constant entropy.
The expansion term, $S$, incorporates local compressibility effects due to heat release from reactions, compositional changes, and external sources,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc,\label{eq:S}
\end{equation}
where $p_{X_k} \equiv \left. \partial p / \partial X_k
\right|_{\rho,T,X_{j,j\ne k}}$, $\xi_k \equiv \left. \partial h /
\partial X_k \right |_{p,T,X_{j,j\ne k}},
p_\rho \equiv \left.
\partial p/\partial \rho \right |_{T, X_k}$, and $\sigma \equiv
p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.

>>>>

\section{Governing Equations}\label{sec:equations}
Low Mach number models for reacting flow were originally derived using asymptotic analysis
\citep{rehm1978equations,majda1985derivation} and used in terrestrial combustion applications
\citep{knio1999semi,day2000numerical}.  These models have been extended to nuclear flames
in astrophysical environments using adaptive algorithms in space and time \citep{Bell:2004}.
In Papers I-III, we derived a model and algorithm suitable for stratified astrophysical flow.
In our model, we take the standard equations of reacting, compressible flow, and recast the equation
of state (EOS) as a divergence constraint on the velocity field.
The resulting model is a series of evolution equations for mass, momentum, and energy, subject
to to an additional constraint on velocity discussed below:
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc.\label{eq:enthalpy}
\end{eqnarray}
Here $\rho$, $\Ub$, and $h$ are the mass density,
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$.  The species are constrained
such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
Here $\Hnuc$ is the nuclear energy generation rate per unit mass.
The pressure is decomposed into a hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that 
$|\pi|/p_0 = \mathcal{O}(M^2)$ (we use $\xb$ to represent the Cartesian coordinate 
directions of the full state and $r$ to represent the radial coordinate direction for 
the base state).  We also define a base state density, $\rho_0 = \rho_0(r,t)$, that represents the lateral average (see Section \ref{Sec:Spatial}) of $\rho$ and is in hydrostatic equilibrium with $p_0$, i.e., 
$\nabla p_0 = -\rho_0 g\eb_r$, where $g=g(r,t)$ is
the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the
outward radial direction.

Mathematically, equations (\ref{eq:momentum})-(\ref{eq:enthalpy}) must still be closed by an EOS.  
This is done by taking the Langrangian derivative of the EOS for pressure as a function of the density, composition, and enthalpy, requiring that the pressure is a prescribed function of altitude and time based on the hydrostatic equilibrium condition.
Details are given in Papers I and II, but to summarize, after substituting the equations for mass and energy into the Lagrangian derivative, the result
is a divergence constraint on the velocity field (\ref{eq:U divergence}),
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).\label{eq:U divergence}
\end{equation}
Here $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho)$ at constant entropy.
The expansion term, $S$, incorporates local compressibility effects due to heat release from reactions, compositional changes, and external sources,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc,\label{eq:S}
\end{equation}
where $p_{X_k} \equiv \left. \partial p / \partial X_k
\right|_{\rho,T,X_{j,j\ne k}}$, $\xi_k \equiv \left. \partial h /
\partial X_k \right |_{p,T,X_{j,j\ne k}},
p_\rho \equiv \left.
\partial p/\partial \rho \right |_{T, X_k}$, and $\sigma \equiv
p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.
To summarize, we model equations (\ref{eq:momentum})-(\ref{eq:enthalpy})
subject to (\ref{eq:U divergence}) and the hydrostatic equilibrium condition.

### 2019-04-30 MAESTROeX_I/intro.tex
paper updates

\section{Introduction} \label{sec:intro}
Many astrophysical flows are highly subsonic; sound waves carry sufficiently little energy that they do not significantly affect the convective dynamics of the system.
In many of these flows, modeling long-time convective dynamics are of interest, and numerical approaches based on compressible hydrodynamics are intractable, even on modern supercomputers.
One approach to this problem is to use low Mach number models.
In a low Mach number approach, sound waves are eliminated from the governing equations while retaining compressibilitiy effects due to, e.g., nuclear energy release, compositional changes, and thermal diffusion.
The resulting model can be numerically integrated with much larger time steps than a compressible model.
Low Mach number models have been developed for a variety of contexts including combustion \citep{day2000numerical}, terrestrial atmospheric modeling \citep{duarte2015low},
and elastic solids \citep{abbate2017all}.

In recent years, a number of alternative approaches to modelling low Mach number flows have been developed. These include semi-implicit all-Mach number solvers, where the Euler equations
are split into an acoustic part and an advective part \citep{Kwatra2009,Degond2009,Cordier2012,Haack2012,Happenhofer2013,Chalons2016,Padioleau2019}. The fast acoustic waves are then
solved using implicit time integration, while the slow material waves are solved explicitly. Another approach is to use preconditioned all-Mach number solvers
\citep{Miczek2014,Barsukow2016}, where the numerical flux is multiplied by a preconditioning matrix. This reduces the stiffness of the system at low Mach numbers, while retaining the
correct scaling behavior. In the reduced speed of sound technique (RSST) and related methods, the speed of sound is artificially reduced by including a suitable scaling factor in the
continuity equation, reducing the restriction on the size of the timestep \citep{Rempel2005,Hotta2012,Wang2015,Takeyama2017,Iijima2018}. The MUSIC code uses fully implicit time integration for the compressible Euler equations, which therefore allows for arbitrarily large timesteps \citep{Viallet2011,Viallet2015,Goffrey2016}.

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
The low Mach number model in MAESTRO is unique in that it is specifically designed for astrophysical settings with significant atmospheric stratification.
Central to the algorithm is a stratified background (or base) state density and pressure held in hydrostatic equilibrium, and also vary as a function of altitude and time.
MAESTRO is a structured-grid, finite-volume code that utilizes adaptive mesh refinement (AMR) to refine grids locally in space.
This code was developed in the pure-Fortran90 FBoxLib framework (see the FBoxLib repository in \cite{AMReX}) for block-structured AMR applications.
MAESTRO is suitable for full spherical stars, as well as planar simulations of dynamics within localized regions of a star.

@@

We also present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising accuracy.
The key design decision was to replace the predicted evolution of the base state with
a predictor-corrector approach.  Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.  Second, we now provide additional spatial mapping
procedures that reduces, and in some cases eliminates, mapping error from the one-dimensional
base state to the Cartesian grid state for spherical problems.
Finally, the code has been completely ported to a new software framework suitable for modern manycore supercomputers.
The resulting MAESTROeX is implemented in the C++/F90 AMReX public software library \cite{AMReX}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub at {\url https://github.com/AMReX-Astro/MAESTROeX} and uses the microphysics
libraries at {\url https://github.com/starkiller-astro/Microphysics} and the AMReX software library for block-structured adaptive mesh refinement
{\url https://github.com/AMReX-Codes/amrex}.

>>>>

\section{Introduction} \label{sec:intro}
Many astrophysical flows are highly subsonic; sound waves carry sufficiently little energy that they do not significantly affect the convective dynamics of the system.
In many of these flows, modeling long-time convective dynamics are of interest, and numerical approaches based on compressible hydrodynamics are intractable, even on modern supercomputers.
One approach to this problem is to use low Mach number models.
In a low Mach number approach, sound waves are eliminated from the governing equations while retaining compressibilitiy effects due to, e.g., nuclear energy release, compositional changes, and thermal diffusion.
The resulting model can be numerically integrated with much larger time steps than a compressible model.

Low Mach number models have been developed for a variety of contexts including combustion \citep{day2000numerical}, terrestrial atmospheric modeling \citep{duarte2015low},
and elastic solids \citep{abbate2017all}.
In recent years, a number of alternative approaches to modeling low Mach number astrophysical flows have been developed.
These include semi-implicit all-Mach number solvers, where the Euler equations are split into an acoustic part and an advective part \citep{Kwatra2009,Degond2009,Cordier2012,Haack2012,Happenhofer2013,Chalons2016,Padioleau2019}.
The fast acoustic waves are then solved using implicit time integration, while the slow material waves are solved explicitly.
Another approach is to use preconditioned all-Mach number solvers \citep{Miczek2014,Barsukow2016}, where the numerical flux is multiplied by a preconditioning matrix.
This reduces the stiffness of the system at low Mach numbers, while retaining the correct scaling behavior. In the reduced speed of sound technique (RSST) and related methods, the speed of sound is artificially reduced by including a suitable scaling factor in the continuity equation, reducing the restriction on the size of the timestep \citep{Rempel2005,Hotta2012,Wang2015,Takeyama2017,Iijima2018}.
The MUSIC code uses fully implicit time integration for the compressible Euler equations, which therefore allows for arbitrarily large timesteps \citep{Viallet2011,Viallet2015,Goffrey2016}.

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for flows where the Mach number is small ($\sim 0.1$ or smaller).  
Furthermore, the low Mach number model in MAESTRO is specifically designed for astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized regions of a star.
MAESTRO uses an explicit Godunov approach for advection, the stiff ODE VODE package \citep{vode} for reactions, and multigrid-based linear solvers for the pressure-projection steps, and is thus time-step limited by an advective CFL constraint based on the fluid velocity, not the sound speed.
Central to the algorithm is a stratified background (or base) state density and pressure held in hydrostatic equilibrium, and also vary as a function of altitude and time.
MAESTRO is a structured-grid, finite-volume code that utilizes adaptive mesh refinement (AMR) to refine grids locally in space.
This code was developed in the pure-Fortran90 FBoxLib framework (see the FBoxLib repository in \cite{AMReX}) for block-structured AMR applications.

@@

We also present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising accuracy.
The key design decision was to replace the predicted evolution of the base state with
a predictor-corrector approach.  Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.  In particular, schemes based on spectral deferred corrections \citep{dutt2000spectral} have been used to generate high-order temporal integrators for problems of reactive flow \citep{dutt2000spectral} and low Mach number combustion \citep{pazner2016high,nonaka2018conservative}
Second, we now provide additional spatial mapping
procedures that reduces, and in some cases eliminates, mapping error from the one-dimensional
base state to the Cartesian grid state for spherical problems.
Finally, the code has been completely ported to a new software framework suitable for modern manycore supercomputers.
The resulting MAESTROeX is implemented in the C++/F90 AMReX public software library \cite{AMReX}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub ({\tt https://github.com/AMReX-Astro/MAESTROeX}) and uses the Starkiller-Astro microphysics libraries ({\tt https://github.com/starkiller-astro/Microphysics}) and the AMReX software library for block-structured adaptive mesh refinement ({\tt https://github.com/AMReX-Codes/amrex}).

The rest of this paper is organized as follows.
In Section \ref{sec:equations} we review our model for stratified low Mach number astrophysical flow.
In Section \ref{eq:algorithm} we present our numerical algorithm in detail.
In Section \ref{sec:results} we validate our new approach and examine the performance of our algorithm on full spherical star problems used in previous scientific investigations.
Finally, we conclude in Section \ref{sec:conclusions}.

### 2019-04-30 MAESTROeX_I/paper.tex
more updates
paper updates

@@

%% Mark off the abstract in the ``abstract'' environment.
\begin{abstract}
We present MAESTROeX, a massively parallel solver for low Mach number astrophysical flows.
Our model equations allow for long-time integration for highly subsonic flows compared to compressible approaches.
The code leverages the new AMReX framework for block-structured adaptive mesh refinement calculations, and also provides additional algorithmic features from the original MAESTRO code.
We present a new temporal integration option that is much simpler than our previous approach while retaining the same order order of accuracy.
We also provide a more accurate spatial discretization for mapping the base state onto the full state Cartesian grid.
Using our previous studies on the convective phase of single-degenerate progenitor models of Type Ia supernovae, we characterize the performance of the code and validate the new algorithmic features.
\end{abstract}

>>>>

\usepackage{url}

\usepackage{mathtools}

@@

%% Mark off the abstract in the ``abstract'' environment.
\begin{abstract}
We present MAESTROeX, a massively parallel solver for low Mach number astrophysical flows.
Our model equations allow for long-time integration for highly subsonic flows compared to compressible approaches.
The code leverages the new AMReX framework for block-structured adaptive mesh refinement calculations allowing for improved scalability of the predecessor code.
We present a new, simplified temporal integration scheme that retains the same order of accuracy as our previous code.
We also explore the use of alternative spatial mapping of the one-dimensional base state onto the full Cartesian grid.
Using our previous studies on the convective phase of single-degenerate progenitor models of Type Ia supernovae, we characterize the performance of the code and validate the new algorithmic features.
\end{abstract}
